---
title: "Midterm 1 W24"
author: "Eva Moncada"
date: "`r Sys.Date()`"
output:
  html_document: 
    keep_md: yes
  pdf_document: default
---

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your code must be organized, clean, and run free from errors. Remember, you must remove the `#` for any included code chunks to run. Be sure to add your name to the author header above. 

Your code must knit in order to be considered. If you are stuck and cannot answer a question, then comment out your code and knit the document. You may use your notes, labs, and homework to help you complete this exam. Do not use any other resources- including AI assistance.  

Don't forget to answer any questions that are asked in the prompt!  

Be sure to push your completed midterm to your repository. This exam is worth 30 points.  

## Background
In the data folder, you will find data related to a study on wolf mortality collected by the National Park Service. You should start by reading the `README_NPSwolfdata.pdf` file. This will provide an abstract of the study and an explanation of variables.  

The data are from: Cassidy, Kira et al. (2022). Gray wolf packs and human-caused wolf mortality. [Dryad](https://doi.org/10.5061/dryad.mkkwh713f). 

## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```

## Load the wolves data
In these data, the authors used `NULL` to represent missing values. I am correcting this for you below and using `janitor` to clean the column names.
```{r message=FALSE, warning=FALSE}
wolves <- read.csv("data/NPS_wolfmortalitydata.csv", na = c("NULL")) %>% clean_names()
```

## Questions
Problem 1. (1 point) Let's start with some data exploration. What are the variable (column) names?  

```{r}
names(wolves)
```

Problem 2. (1 point) Use the function of your choice to summarize the data and get an idea of its structure.  

```{r}
summary(wolves)
```

Problem 3. (3 points) Which parks/ reserves are represented in the data? Don't just use the abstract, pull this information from the data.  

```{r}
tabyl(wolves, park)
```

Problem 4. (4 points) Which park has the largest number of wolf packs?

```{r}
wolves %>% 
  select(park, packsize_aug) %>% 
  group_by(park) %>% 
  arrange(desc(packsize_aug))
```

Problem 5. (4 points) Which park has the highest total number of human-caused mortalities `mort_all`?

```{r}
wolves %>% 
  select(park, mort_all) %>% 
  group_by(park) %>% 
  arrange(desc(mort_all))
```

The wolves in [Yellowstone National Park](https://www.nps.gov/yell/learn/nature/wolf-restoration.htm) are an incredible conservation success story. Let's focus our attention on this park.  

Problem 6. (2 points) Create a new object "ynp" that only includes the data from Yellowstone National Park. 

```{r}
ynp <- filter(wolves, park=="YNP")
```



Problem 7. (3 points) Among the Yellowstone wolf packs, the [Druid Peak Pack](https://www.pbs.org/wnet/nature/in-the-valley-of-the-wolves-the-druid-wolf-pack-story/209/) is one of most famous. What was the average pack size of this pack for the years represented in the data?
```{r}
anyNA(ynp$packsize_aug)
```

```{r}
  mean(ynp$packsize_aug, na.rm=T)
```


Problem 8. (4 points) Pack dynamics can be hard to predict- even for strong packs like the Druid Peak pack. At which year did the Druid Peak pack have the largest pack size? What do you think happened in 2010?

```{r}
ynp %>% 
  filter(pack=="druid") %>% 
  select(biolyr, packsize_aug) %>% 
  group_by(biolyr) %>% 
  arrange(desc(packsize_aug))
```

```{r}
ynp %>% 
  filter(pack=="druid") %>% 
  select(biolyr, packsize_aug) %>% 
  group_by(biolyr) %>% 
  arrange(desc(biolyr))
```

In 2010, the Druid Peak wolf pack went extinct. The pack size was a mere 12 wolves in 2009. These wolves could have been killed due to hunters or possibly their habitat was destroyed. Another competing animal could have been introduced into the park, where competing for similar foods could have ruined their chances of being able to feed themselves. Since there are other wolf packs in the park, they could have been competing with one another, leading to the decline of the Druids Peak wolfpack.

Problem 9. (5 points) Among the YNP wolf packs, which one has had the highest overall persistence `persisty1` for the years represented in the data? Look this pack up online and tell me what is unique about its behavior- specifically, what prey animals does this pack specialize on?  

```{r}
ynp %>%
  select(pack, persisty1, biolyr) %>% 
  group_by(pack) %>%
  summarize(total_persistence = sum(persisty1)) %>%
  arrange(desc(total_persistence))  
```

The mollies wolf pack had the highest overall persistence amongst the years in the data. According to Google, this wolf pack preys on large animals, specifically elk and bison. This wolf pack was not only the first in Yellowstone to eat bison, but was also one of the longest lasting wolf packs at Yellowstone. What is unique about their behavior is that there are many strong "alpha-females". the carcasses from the bison they hunt are used to feed other animals in the park, including Grizzly Bears.


Problem 10. (3 points) Perform one analysis or exploration of your choice on the `wolves` data. Your answer needs to include at least two lines of code and not be a summary function.  


```{r}
wolves %>% 
  filter(pack=="Pinto") %>% 
  select(pack, biolyr, reprody1) %>% 
  arrange(desc(biolyr))
```
For this last question, I was curious on if the Pinto pack had reproduced and wanted to arrange it by year. According to the wolves data, the Pinto pack only reproduced in the year 2007. Pup were observed in 2007 but not in 2006. 
