---
title: "Lab 4 Homework"
author: "Eva Moncada"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the tidyverse
```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

## Data
For the homework, we will use data about vertebrate home range sizes. The data are in the class folder, but the reference is below.  

**Database of vertebrate home range sizes.**  
Reference: Tamburello N, Cote IM, Dulvy NK (2015) Energy and the scaling of animal space use. The American Naturalist 186(2):196-211. http://dx.doi.org/10.1086/682070.  
Data: http://datadryad.org/resource/doi:10.5061/dryad.q5j65/1  

**1. Load the data into a new object called `homerange`.**
```{r}
homerange <- readr::read_csv ("data/Tamburelloetal_HomeRangeDatabase.csv")
```
```{r}
homerange
```

**2. Explore the data. Show the dimensions, column names, classes for each variable, and a statistical summary. Keep these as separate code chunks.**  
```{r}
dim(homerange)
```

```{r}
names(homerange)
```

```{r}
homerange$class
```

```{r}
summary(homerange)
```

**3. Change the class of the variables `taxon` and `order` to factors and display their levels.**  
```{r}
homerange$taxon <- as.factor(homerange$taxon)
homerange$order <- as.factor(homerange$order)
```

```{r}
homerange$taxon
```


**4. What taxa are represented in the `homerange` data frame? Make a new data frame `taxa` that is restricted to taxon, common name, class, order, family, genus, species.**  

Changing some of the names since they incorrectly written. 

```{r}
colnames(homerange) <- c("taxon", "common_name", "class", "order", "family", "genus", "species", "primary_method", "N", "mean_mass_g", "log10_mass", "alternative_mass_reference", "mean_hra_m2", "log10_hra", "hra_reference", "realm", "thermoregulation", "locomotion", "trophic_guild", "dimension", "prey_mass", "log10_prey_mass", "PPMR", "prey_size_reference")
```


```{r}
taxa <- homerange[, c("taxon", "common_name", "class", "order", "family", "genus", "species")]
head(taxa)
```

**5. The variable `taxon` identifies the common name groups of the species represented in `homerange`. Make a table the shows the counts for each of these `taxon`.**  
```{r}
table(homerange$taxon)
```

**6. The species in `homerange` are also classified into trophic guilds. How many species are represented in each trophic guild.**  
```{r}
table(homerange$trophic_guild)
```

**7. Make two new data frames, one which is restricted to carnivores and another that is restricted to herbivores.**  

```{r}
carnivores <- filter(homerange, trophic_guild == "carnivore")
herbivores <- filter(homerange, trophic_guild == "herbivore")
```

**8. Do herbivores or carnivores have, on average, a larger `mean.hra.m2`? Remove any NAs from the data.**  
```{r}
anyNA(carnivores$mean_hra_m2)
```

```{r}
mean(carnivores$mean_hra_m2, na.rm =T)
```


```{r}
anyNA(herbivores$mean_hra_m2)
```

```{r}
mean(herbivores$mean_hra_m2, na.rm =T)
```


**9. Make a new dataframe `owls` that is limited to the mean mass, log10 mass, family, genus, and species of owls in the database. Which is the smallest owl? What is its common name? Do a little bit of searching online to see what you can learn about this species and provide a link below** 

```{r}
homerange$order
```

```{r}
homerange %>% 
  select(taxon, order, mean_mass_g, log10_mass, family, genus, species) %>% 
  filter(taxon=="birds", order== "strigiformes") %>% 
  arrange(desc(mean_mass_g))
```
The smallest owl is species passerinum from the glaucidium, genus family (Glaucidium passerinum). This  owl is the smallest in Europe with a wingspan of 32-39 centimeters and weight of only 47-83 grams,which is equivalent to around 0.10-0.18 pounds. This owl is also found in areas of Asia including Mongolia and China. This species lives in coniferous forests and mountainous areas. These owls do not migrate and hence hoard their carnivorous foods in the autumn to be consumed in the winter. Their carnivorous diet primarily consists of small mammals, small birds, lizards, fish and insects. 

Source: https://animalia.bio/eurasian-pygmy-owl?letter=e

**10. As measured by the data, which bird species has the largest homerange? Show all of your work, please. Look this species up online and tell me about it!**.  
```{r}
homerange %>% 
  arrange(desc(mean_hra_m2)) %>% 
  filter(class == "aves")
```
The cheriway species, or Caracara, has the largest homerange. The Crested Caracara are omnivorous falcons that primarily live in Central and Southern Americas, along with some southern states in the United States. These birds are black and white with yellow-orange skin on their legs and around their bills.They are medium sized with their wingspan of 122-125 centimeters and lengths of 49-58 centimeters. This species is the only falcon that build their own nests while other falcons use old nests created by other species. 

Source: https://www.allaboutbirds.org/guide/Crested_Caracara/overview 

## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences.   