---
title: "lab_4_warm_up"
author: "Eva Moncada"
date: "2024-01-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Load the libraries
You need tidyverse and palmer penguins. load bot using library().
 
```{r}
library(tidyverse)
library(palmerpenguins)
```
 
```{r}
penguins
```
 
## 2. What are the dimensions of the "penguins" dataset?
 
```{r}
dim(penguins)
```
 
## 3. hat are the names of the variables? Variables are the COLUMNS of the data set. 
 
```{r}
names(penguins)
```

You can also do:

```{r}
glimpse(penguins)
```

## 4. How many individuals were sampled on each island?
We are making the data set an object. So we can now print out the data to see it.
```{r}
penguins <- penguins
```

```{r}
penguins
```

Now we can use the function Table(). The $ sign allows you to choose a variable within that dataset. 

```{r}
table(penguins$island)
```
Or you can do:
```{r}
summary(penguins$island)
```

```{r}
mean(penguins$body_mass_g)
```

Notice hw it responds with NA. This is because there are some NA'sin the data. so we use na.rm=T t repve the NAs/
```{r}
mean(penguins$body_mass_g, na.rm=T)
```
